Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics

### 2013-10-01 paper/paper2.tex
paper changes. Wrote out most of RTM's notes into paragraph form. The most major thing we need is a discussion of why this is better than Liz's method. @rmcgibbo I think you said you had an idea of what to write here? Other than that we just need to make small changes to make it flow a bit better.

\usepackage{amsmath}    % need for subequations
\usepackage{amssymb}    % for symbols
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
\begin{document}

@@

\section{Introduction}
Protein dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, the conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics occurring via various pathways and with multiple intermediates \cite{Cosa2006Evidence, Zhang2011Direct}. But atomic-resolution structural characterization of these dynamics is often a Herculean challenge, particularly in dynamical settings -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs/mull_ww.png}
\caption{Three systems studied in this work. (a) Langevin dynamics on the two dimensional M\"{u}ller potential. (b) 50 ns of dynamics of alanine dipeptide. (c) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\end{figure}

@@

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated [HOW MANY STEPS?] of Langevin dynamics on the M\"{u}ller potential, clustered the data using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with [HOW MANY ROUNDS] of Monte carlo rejection sampling, using a padding distance of $R=?$ around the state centers from the 500 state model to define the datasets envelope.

\begin{figure}[h]
\centering
\includegraphics[width=3.3in]{figs/mull_cross.png}
\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states, very close to the region in which the BIC is minimized. The BIC optimal model is also where the models' slowest implied timescals begins to become invariant with respect to the number of states.}
\label{fig:mullerlike}
\end{figure}

As shown in Fig. \ref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly. Here, the BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate. This model selection is consistent with tests based on the invariance of transition matrix eigenspectrum, but permits a direct optimization of the state space.

\begin{figure}
\centering
\includegraphics[width=2in]{figs/ww_aic_vs_time.png}
\caption{Implied timescales and BIC scores for a series of Markov state models for the Fip35 WW domain. The panels differ by the dimensionality of the tICA projection. All models were built using Markovian lag time of [HOW MANY ns].}
\label{fig:ww}
\end{figure}

\subsection{Alanine Dipeptide}

@@

To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic}. In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis\cite{Schwantes2013Improvements}, retaining between two and six uncorrelated(?) linear combination of pairwise distances(?).

How many states are required for a Markov state model? Lane et. al., when analyzing this same dataset, proposed a 26,000 state Markov state model after clustering the data by RMSD, whereas Kellogg et. al. using an approach similar to that described herein, arrive at a [HOW MANY] state model after clustering on contact maps\cite{Lane2011Markov, Kellogg2012Evaluation}.

\begin{itemize}
\item Is the 26,000 state model ``overfit''? The obvious answer, based on the line of reasoning that this paper is taking, is yes. But I think we need to consider the question a little more subtly. Perhaps it depends on what questions we ask of the model. When you ask both a ``properly-fit'' model and an overfit model a question that basically falls within the training set -- and neither of them need to extrapolate or really infer out-of-sample information, they'll both do fine. Heck, the overfit model might do better at telling you exactly what was \emph{in} the dataset.
\item The folding timescale is relatively insensitive to differences in MSM construction methodology. This has been observed in many datasets (McGibbon 2013, others). We hypothesize that this behavior is due to the large structural deviations at play -- even a two state model, given the same dataset, could basically get the right timescale just from counting the number of transitions. But finding these other timescales that are structurally subtle is different challenge. Also, how robust are other properties like experimental projections or the MFPT distribution to the number of states?
\item Single trajectory datasets are ``easier'' for MSMs. One of the uses of MSMs is to ``stich together'' multiple trajectories -- this function only happens when the states are large enough that they \emph{connect} trajectories, so you can't have too many states in this case. But when you have a single trajectory, there's no possible way to create a disconnected model anyways.
\item We need to state, elegantly if possible, the differences between our likelihood and Liz Kellogg's. This requires some careful language, since we think there are methodological problems with her approach.
\end{itemize}

@@

@@

\section{Conclusions}
\begin{itemize}
    \item MSMs are great
    \item They have weaknesses. We are solving them.
    \item The automated selection of the number of states is a critical part of procedures that build MSMs as part of a simulation protocol to drive adaptive sampling. If MD algorithms required user input every couple picoseconds, people would never get anything done.
\end{itemize}

>>>>

\usepackage{amsmath}    % need for subequations
\usepackage{amssymb}    % for symbols
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
\usepackage{soul}

\begin{document}

@@

\section{Introduction}
Protein dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, the conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with infcreasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics occurring via various pathways and with multiple intermediates \cite{Cosa2006Evidence, Zhang2011Direct}. But atomic-resolution structural characterization of these dynamics is often a Herculean challenge, particularly in dynamical settings -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs/mull_ww.png}
\caption{Three systems studied in this work. (a) Langevin dynamics on the two dimensional M\"{u}ller potential. (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\end{figure}

@@

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated $1\times 10^7$ steps of Langevin dynamics on the M\"{u}ller potential, clustered the data using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with $1\times 10^5$ rounds of Monte carlo rejection sampling, using a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.

The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and furthermore, the likelihood remained almost exactly constant, which gives us some confidence that the volumes do not need to be perfect in order to have a good estimate of the likelihood. \hl{I can make a figure, and we can put it in the SI}.

\begin{figure}[h]
\centering
\includegraphics[width=3.3in]{figs/mull_cross.png}
\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states. There are many methods that can be used to test the transferability of an MSM. We've plotted the AIC, BIC, and cross-validation scores for the WW domain. These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily. \hl{See SI section: ``AIC vs BIC" for a discussion of the various methods in detail.}}
\label{fig:mullerlike}
\end{figure}

As shown in Fig. \ref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly. Here, the AIC and BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate. The optimal models according to the three methods are between 200 and 1000 states for this system. This observation is consistent with the convergence of the timescales of the MSMs, but allows for a direct optimization of the state space, as opposed to a heuristic way of comparing models.

Interestingly, the AIC and BIC penalize the larger state models much more than the cross-validation approach. This could be due to the simplicity of this system. Since the M\"uller potential is only two-dimensional, it is difficult to actually produce a drastically over-fit model. We suspect that the cross-validation score on a real protein system would be closer to the BIC and AIC. However, we note, that cross-validation requires estimating a model without some subset of the data, in which case, it is not feasible for large systems that are not over-sampled.

\begin{figure}
\centering
\includegraphics[width=2in]{figs/ww_aic_vs_time.png}
\caption{Implied timescales and AIC scores for a series of Markov state models for the Fip35 WW domain. The panels differ by the dimensionality of the tICA projection. All models were built using a Markovian lag time of 50 ns. Qualitatively, the optimal model according to the AIC, corresponds roughly to where the timescales begin to flatten out.}
\label{fig:ww}
\end{figure}

@@

To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic}. In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis\cite{Schwantes2013Improvements}, retaining between two and six uncorrelated linear combinations of residue-residue distances.

How many states are required for a Markov state model? Lane et. al., when analyzing this same dataset, proposed a 26,000 state Markov state model after clustering the data by RMSD, whereas Kellogg et. al. using an approach similar to that described herein, arrive at a \hl{1,000} state model after clustering on contact maps\cite{Lane2011Markov, Kellogg2012Evaluation}. \hl{That number is the micro state model from KCenters and CM's, but they also did KMeans and used 100 states}.

According to this likelihood approach, the most likely models built on tICA projections are in the hundreds of states range. This does not mean, however, that the previous models are ``wrong." The value of a model can only be judged based on the questions one is asking. If the question is, what are the slowest dynamics in the data, then an over-fit model may actually give a better estimate than a ``perfectly-fit" model. However, if the question requires an inference of out-of-sample information, then we would trust the well-fit model over an over-fit one.

For protein simulations, the slowest timescale (typically corresponding to folding) has been observed to be insensitive to differences in MSM construction methodology \cite{McGibbon2013Learning, others?}. We hypothesize that this behavior is due to the large structural changes associated with a folding event -- even a two state model, given the same dataset, could estimate the slowest timescale by simply counting the number of transitions. However, teasing out the slow timescales that are associated with subtler structural changes is a different challenge.

\hl{Kellogg stuff.}

@@

The likelihood described above allows for the comparison of MSMs with varying number of states, however it requires the definition of a vector space. As such, models built on different representations cannot be compared. This is because the likelihood requires the calculation of relative volumes, but the volumes will be different depending on what space you are working in. In addition, the lag time must be specified and cannot be selected using the likelihood. This is an improvement over the current state of constructing an MSM where the number of states is selected and then the lag time is tuned according to the eigenspectrum of the model. Selecting a lag time {\it a priori} is a much more intuitive parameter, that one typically knows based on the system.

We can also extend these results to models that do not have discrete states. Since all we require is an emission probability into a vector space, it is easy to use this likelihood for evaluating models that no longer have discrete states.

\section{Conclusions}

Markov State Models are a promising analysis technique for automatically analyzing simulations generated from Markovian dynamics. There are a number of steps in the construction process, however, that require human intervention, which limits the use of MSMs to experts as well as introduces significant uncertainty into the model building process. These results take a large step toward fully automating the construction process by providing a quantitative way of selecting the number of states to use, as well as a way of selecting the clustering method. This will allow for non-experts to begin to use MSMs as well as allow adaptive sampling methods to be applied without human intervention.

\begin{itemize}
\item Is the 26,000 state model ``overfit''? The obvious answer, based on the line of reasoning that this paper is taking, is yes. But I think we need to consider the question a little more subtly. Perhaps it depends on what questions we ask of the model. When you ask both a ``properly-fit'' model and an overfit model a question that basically falls within the training set -- and neither of them need to extrapolate or really infer out-of-sample information, they'll both do fine. Heck, the overfit model might do better at telling you exactly what was \emph{in} the dataset.
\item The folding timescale is relatively insensitive to differences in MSM construction methodology. This has been observed in many datasets (McGibbon 2013, others). We hypothesize that this behavior is due to the large structural deviations at play -- even a two state model, given the same dataset, could basically get the right timescale just from counting the number of transitions. But finding these other timescales that are structurally subtle is different challenge. Also, how robust are other properties like experimental projections or the MFPT distribution to the number of states?
\item Single trajectory datasets are ``easier'' for MSMs. One of the uses of MSMs is to ``stitch together'' multiple trajectories -- this function only happens when the states are large enough that they \emph{connect} trajectories, so you can't have too many states in this case. But when you have a single trajectory, there's no possible way to create a disconnected model anyways.
\item We need to state, elegantly if possible, the differences between our likelihood and Liz Kellogg's. This requires some careful language, since we think there are methodological problems with her approach.
\end{itemize}

@@

\begin{itemize}
    \item MSMs are great
    \item They have weaknesses. We are solving them.
    \item The automated selection of the number of states is a critical part of procedures that build MSMs as part of a simulation protocol to drive adaptive sampling. If MD algorithms required user input every couple picoseconds, people would never get anything done.
\end{itemize}
